Error Bounds for Extended Source Inversion applied to an Acoustic Transmission Inverse Problem

### 2022-01-11 repro/papers/discreptheorypub/paper.tex
more

\title{Solution of an Acoustic Transmission Inverse Problem by
  Extended Inversion: Theory}
\author{William W. Symes\footnotemark[1]}
\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA, {\tt symes@rice.edu}}

\lefthead{Symes}

\righthead{1D Acoustic Inversion: Theory}

\maketitle
\begin{abstract}
  A single-trace transmission inverse problem for the wave equation
  seeks to determine both the wave velocity in a homogenous acoustic
  medium and the transient waveform of an isotropic point source. The
  duration (support) of the source waveform and the source-to-receiver
  distance are assumed known. A least squares formulation of this
  problem exhibits the ``cycle-skipping'' behaviour observed in field
  scale problems of this type, with many local minima differing
  greatly from the global minimizer. This behaviour is eliminated by
  dropping the hard support constraint on the source waveform,
  replacing it by a soft penalty implemented as a weighted mean-square
  of the source waveform. For properly chosen weight function,
  penalizing nonzero values away from $t=0$, the velocity component of
  any stationary point differs from the global minimizer of the
  constrained least-squares formulation by a linear combination of the
  source waveform support radius and data noise-to-signal ratio. Given
  an estimate of data noise, the penalty weight can be dynamically
  adjusteded during iterative optimization to maximize predicted data
  accuracy and closely approximate a support-constrained source waveform.

@@

\section{Introduction}
Inverse problems based on wave equations (acoustic, elastic,
Maxwell's...) are commonplace in geophysics, nondestructive materials
testing, medical imaging, and other areas of science and engineering
in which wave motion plays an important role. Often these problems are
formulated as optimization of mean square (or other scalar measure)
functions of misfit between observed and predicted data. In many
contexts, such as exploration seismology, the computational size of
these problems is very large, and even the prediction (or modeling) of
the data requires use of high-performance computing
environments. Accordingly, rapidly convergent optimization algorithms
based on Newton's method and its relatives to update model parameters
are the numerical methods of choice. These however are local, in the
sense that they converge to stationary points of the objective
functions, usually near the initial model parameters. For
high-frequency (hence high resolution) data regimes, stagnation of
local optimization methods at physically irrelevant model vectors is
frequently observed. This ``cycle-skipping'' is one of the main
obstacles to widespread successful deployment of optimization-based
methods for inverse problems in wave propagation
\cite[]{GauTarVir:86,VirieuxOperto:09,Fichtner:10,Plessix:10,Schuster:17}.

This paper shows how a modification of least-squares data fitting via
{\em modeling operator extension} overcomes cycle-skipping in a simple
example. The problem studied here may be the simplest inverse wave
problem that exhibits cycle-skipping - in fact, essentially this
example is routinely used to illustrate cycle-skipping (see for
example \cite{VirieuxOperto:09}, Figure 7). Its data is a single trace
at a point in $\bR^3$ (``receiver location'') of the causal linear
acoustic pressure field, generated by an isotropic radiator acting at
another point (``source location'') at a known distance from the
receiver location. The acoustic material is supposed spatially
homogeneous, hence characterized by wave slowness (reciprocal
velocity) and density. The object of the inverse problem is to recover
the slowness and the time history of the isotropic radiator
(``wavelet'') from the recorded trace. In order that the data
constrain the slowness, the wavelet is presumed to have known compact
support. Thus the (``physical'') modeling operator has as its domain
the Cartesian product of an interval of positive slownesses, and the
subspace of $L^2(\bR)$ consisting of wavelets with the prescribed
support. Even for noise-free data (in the range of the modeling
operator), the obvious least squares problem formulated in terms of
this modeling operator exhibits many local minimizers having nothing
to do with the (``target'') slowness and wavelet used to create the
data - that is, cycle-skipping occurs (Theorem \ref{thm:fwi}).

Dropping the support constraint on the wavelet creates an extension of
the modeling map for which any data may be fit exactly, for any choice
of slowness. The link between data and slowness is restored by adding
a quadratic penalty on non-zero values of the wavelet outside of the
physical support interval to the extended mean-square data misfit. The
first main result of this paper is that for noise-free data, the
slowness component of {\em any} stationary point of this extended
penalty function differs from the target slowness by a multiple of the
target wavelet radius (Theorem \ref{thm:rampreallygood}). That is, cycle-skipping does not occur, and the
wavelet support radius determines the resolution of slowness. For
noisy data (that is, sum of a noise-free signal trace in the range of
the physical modeling operator and a square-integrable noise trace),
any stationary point differs from the target slowness by a combination
of the wavelet support radius and the noise-to-signal ratio (Theorem \ref{thm:mnoiseres}). Since the
wavelet component of the extended domain does not constrain support,
stationary points of the extended penalty function do not provide
solutions of the original, physical inverse problem, which includes
the support constraint. Such solutions can be obtained, with estimates
on support radius and data error, simply by truncating the wavelet
components of stationary points (Theorem \ref{thm:ipnoisesuf}).

The inverse problem studied here is far too simple to have any direct
use in applications. However it seems worthy of careful study for two
reasons. First, its simplicity allows an unusually complete account of
its properties. The present paper details its mathematical behaviour,
including explicit and sharp error estimates rarely available for nonlinear
inverse problems. A companion paper \cite[]{SymesChenMinkoff:21}
describes numerical examples that illustrate the mathematical
conclusions.

Second, the analysis reveals several important features shared with
similar approaches to other wave inverse problems, and so provides an
at least partial pattern for reaching similar conclusions with
immediate practical implications. For example, nested optimization
(variable projection method, \cite{GolubPereyra:03}) based on a model
decomposition into inner and outer variables seems to be very
important: in some cases, such as the simple problem presented here,
the decomposition is obvious (inner variable = wavelet, outer variable
= slowness), in others less so
\cite[]{geoprosp:2008,Terentyev:thesis}. The extended modeling
operator must be surjective, or at least have a dense range, {\em for
  each value of the outer variable}, so that data may be fit well even
for a poor initial guess of the outer variable. for the extended
inversion approach to be successful. The derivative of the extended
modeling operator with respect to the outer variable is
well-approximated by the composition of the extended modeling operator
itself and a pseudodifferential operator of order 1
\cite[]{Symes:IPTA14,tenKroode:IPTA14}. The choice of the penalty
operator, controlling the extended degrees of freedom, is critical: in
order to produce an objective immune from cycle-skipping, this penalty
operator must be (pseudo-)differential
\cite[]{StolkSymes:03}. Generally, the extended modeling operator is
not differentiable in the outer variable. However these last two facts
together imply the differentiability of the reduced objective function
from variable projection (Theorem \ref{thm:diffobj}, Appendix C). Because of the data-fitting
assumption, a straightforward algorithm for scaling the quadratic
penalty, based on the Discrepancy Principle, is available - see
\cite{FuSymes2017discrepancy,SymesChenMinkoff:21}, also Appendix
A. This scaling varies dynamically during iterative optimization, in
effect changing the objective function sporadically as the iteration
converges.

The approach explained here is an example of {\em extended source}
inversion, see \cite{HuangNammourSymesDollizal:SEG19} for a review,
and \cite{MetivierBrossier:SEG20} for a recent variation. Adding
degrees of freedom to the coefficients in the wave equation in various
ways produces so-called {\em extended model} inversion approaches,
which actually have a longer history in seismology - see
\cite{geoprosp:2008} for details. The need to overcome cycle-skipping
has spawned a great number of concepts, of which extension-based
inversion is only one. Least-squares inversion succeeds with
sufficiently good initial estimates of wave velocity, and the dominant
techniques in industrial and academic seismology exploit data other
than direct measurements of waves, for example measurement of wave
time-of-flight, to supply such initial estimates
\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. Use of a data
misfit measure other than least squares, such as versions of the
Wasserstein metric from optimal transport theory, also shows promise
\cite[]{Metivier:GEO18,EngquistYang:GEO18}.

The next section gives precise statements of the inverse problem
studied here, and its various components. The following sections
contain statements and proofs of the main results: the existence of
muliple stationary points of the least squares problem, properties of
the extended modeling operator and the reduced objective, choice of
the penalty operator, the main results mentioned above about
stationary points of the reduced objective, and the discrepancy-based
algorithm for adjusting the penalty scale. Three appendices justify
the discrepancy-based scale adjustment, review the link between
constrained support and spectrum provided by the Heisenberg inequality,
and expose the abstract structure underlying the differentiability of
the reduced objective.

I use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
the algebra of bounded linear operators from the Hilbert space $X$ to 
the Hilbert space $Y$, and its subalgebra of invertible
operators. All of the operators appearing in this discussion are
members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
operator in the sense specified by its domain and range Hilbert structure.

\section{An Acoustic transmission Inverse Problem}
According to linear acoustics, the causal pressure field due to an
isotropic point radiator at $\bx_s \in \bR^3$ 
solves the the wave equation \cite[]{Frie:58}:
\begin{eqnarray}
  \label{eqn:awe}
  \left(m^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
  p(\bx,t)&=&0, t\ll 0.
\end{eqnarray}
The solution is well-known, see for instance
\cite{CourHil:62}, Chapter VI, section 12, equation 47:
\begin{equation}
  \label{eqn:homsol}
  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right),
\end{equation}
This trace of $p$ at $\bx_r \in \bR^3$ can be viewed as the result of
applying an m-dependent linear operator $F[m]$ to the wavelet $w$:
\begin{equation}
\label{eqn:mod}
F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right) 
\end{equation}
The slowness $m$ must be positive, as follows from basic acoustics,
and in fact reside in a range characteristic of the
material model: for crustal rock, a reasonable choice would be
$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.

Natural choices for domain and
range of $F$ are thus
\begin{itemize}
\item $M=(m_{\rm min}, m_{\rm max}),\,0 < m_{\rm min} \le m_{\rm
    max}$;
\item $W = L^2(\bR)$;
\item $D=L^2([t_{\rm min},t_{\rm max}]),\, t_{\rm min}<t_{\rm max}$;
\item $F: M \times W \rightarrow D$ as specified in \ref{eqn:mod}.
\end{itemize}
It is immediately evident from these choices and from the definition
\ref{eqn:mod} that
\begin{equation}
  \label{eqn:mapprop}
  \mbox{for }m \in M, F[m] \in {\cal B}(W,D),\mbox{ and }\|F[m]\| =
  \frac{1}{4\pi r}.
\end{equation}
Note also that $F[m]$ is surjective for every $m \in M$.

\noindent{\bf Remark:} In computational practice, $W$ will have to be replaced by a
finite-dimensional subspace of $L^2(\bR)$. Many such choices will
implicitly limit the support of $w \in W$ to a bounded interval, say
$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
\begin{eqnarray}
  \label{eqn:dombounds}
  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
\end{eqnarray}
I will ignore these computational necessities in this work,
maintaining the definition $W=L^2(\bR)$.

As mentioned earlier, $F[m]$ is surjective for every $m \in M$. Since
all possible data lie in the range of $F[m]$ for any $m \in M$, some
restriction of the domain of $F$ is necessary in order that fitting
the data constrain $m$. The constraint employed is the specification
of a maxium support radius $\lambda_{\rm max} >0$. Then define for
$\lambda \in (0,\lambda_{\rm max}]$:
\begin{itemize}
\item $\lW = \{w \in W:
  \mbox{supp }w \subset [-\lambda,\lambda]\}$;
\item $\lF = F|_{M \times \lW}$.
\end{itemize}

@@

\noindent{\bf Remark:} The constraint $\epsilon < 1$ imposed on the
target noise level is eliminates the obvious choice $(m,0)$, which
satisfies the data misfit constraint for any $m \in M$.

@@

\noindent{\bf Remark:} The support constraint is closely linked to the
folk theorem about FWI noted many times in the literature: convergence
of a descent method requires that the initial slowness must be known
to ``within a (fraction of a) wavelength''. The relation is a
consequence of Heisenberg's inequality, and will be reviewed in
Appendix B.

@@

\section{Full Waveform Inversion}
While $F$ is surjective, as noted above, it is
very far from injective. On the other hand, under a constraint that
will be assumed throughout, $\lF[m]$ is injective for each $m \in M$ (in fact, $4 \pi r
\lF[m]$ is an isometry):
\begin{proposition}
  \label{thm:fullrec}
  Suppose that 
  \begin{equation}
    \label{eqn:fullrec}
    [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
    \subset [t_{\rm min},t_{\rm max}].
  \end{equation}
  Then $\lF[m]\in {\cal B}(\lW,D)$ is coercive for every $m \in M, \lambda \in
  (0,\lambda_{\rm max}]$.
\end{proposition}

@@

\begin{proof} From the definition \ref{eqn:mod},
\[
 \lerr[m,w;d] =  \frac{1}{32\pi^2
    r^2\|d\|^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
\]
Since $w_*, w$ vanish for $|t|>\lambda$,
$\lF[m_*]w_*(t)$ vanishes if $|t-m_*r|>\lambda$ and $\lF[m]w$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
either $\lF[m]w(t)=0$ or $\lF[m_*]w_*=0$. Therefore $\lF[m]w$ and
$\lF[m_*]w_*$ are orthogonal in the sense of the $L^2$ inner product
$\langle \cdot,\cdot \rangle_D$ on $D$:
\begin{equation}
  \label{eqn:ortho}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
  F[m_*]w_*\rangle_D = 0
\end{equation}
But $d = \lF[m_*]w_*$, so this is the same as saying that $d$ is
orthogonal to $F[m]w$. So conclude after a minor manipulation that
\[
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
\]
\begin{equation}
  \label{eqn:iso}
  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
\end{equation}
That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
$m$, and its minimum over $w$ is attained for $w=0$
\end{proof}

Therefore local minimizers of $\lerr$ abound, as far as you like from the
global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
$e$ gives no useful information whatever about constructive search
directions, and descent-based optimization tends to fail if the
initial estimate $m_0$ is in error by more than $2\lambda/r$
(``further than a multiple of a wavelength'', per discussion
in the second Appendix). In fact the actual behaviour of FWI itererations is worse
(failure if $m_0$ is in error by ``half a wavelength''), as follows
from a more refined analysis of the cycle-skipping local behaviour of $\lerr$ near its
global minimizer.

@@

\section{Variable Projection}
The main theoretical device used in the proofs of our main results on 
extended inversion is Variable Projection reduction of the penalty objective $\Ja$
(equation \ref{eqn:penerr}) to a function $\tJa$ of 
$m$ alone, by minimization over $w$:
\begin{equation}
  \label{eqn:redexp}
  \tJa[m;d] = \inf_w \Ja[m,w;d]
\end{equation}

A minimizer $w$ on the right-hand side of definition
\ref{eqn:redexp} must solve the {\em normal equation}
\begin{equation}
  \label{eqn:norm}
  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
\end{equation}

@@

With these choices, the normal equation \ref{eqn:norm1} becomes
\begin{equation}
\label{eqn:norm2}
\left(\frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
      t_{\rm max}-mr]} + \alpha^2a^2\right)w= F[m]^Td.
\end{equation}

\begin{proposition}
  \label{thm:norminvexp}
  Assume the conditions \ref{eqn:mod}, \ref{eqn:fullrec},
  \ref{eqn:annmult}. Also assume that $\lambda \in (0,\lambda_{\rm
    max}], \alpha > 0,$ and that  $C>0$ exists so that $a \in L^{\infty}(\bR)$
  mentioned in condition \ref{eqn:annmult} satisfies condition
  \ref{eqn:abnd}.
%  \begin{equation}
%   \label{eqn:abnd}
%    |t| > \lambda \Rightarrow a(t) \ge C,
%  \end{equation}
%  in which $C>0$ may depend on $\lambda$.
  Then
  \begin{itemize}
  \item[1. ]the normal operator $F[m]^TF[m] + \alpha^2A^TA$ is
    invertible for any $m \in M$, $\alpha > 0$;
  \item[2. ]the solution $\aw[m;d]\in W$ of the normal equation
    \ref{eqn:norm} is given by
    \begin{equation}
      \label{eqn:normsol}
      \aw[m;d](t) = \left\{
        \begin{array}{c}
          \left(\frac{1}{(4\pi r)^2} + \alpha^2
          a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
          min}-mr, t_{\rm max}-mr];\\
          0, \mbox{ else;}
        \end{array}
      \right.
    \end{equation}
  \item[3. ]if in addition $d=F[m_*]w_*, w_* \in \lW$ is noise-free, as in equation
    \ref{eqn:defdatanonoise},
    \begin{equation}
      \label{eqn:solnonoise}
      \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right).
    \end{equation}
  \end{itemize}
\end{proposition}

@@

\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
were differentiable, then ``local minimizer'' in the conclusion of the
preceding theorem could be replaced by ``stationary point''. However,
for the problem addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is
  not} differentiable without added smoothness constraints on $w$,
whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
penalty operator $A$.

>>>>

\title{Error Bounds for Extended Source Inversion applied to an Acoustic Transmission Inverse Problem}
\author[Symes]{William W. Symes$^1$\\
$^1$ PO Box 43, Orcas, WA 98280 USA, {\tt symes@rice.edu}}

\maketitle
\begin{abstract}
  A simple inverse problem for the wave equation requires
  determination of both the wave velocity in a homogenous acoustic
  material and the transient waveform of an isotropic point radiator,
  given the time history of the wavefield at a remote point in
  space. The duration (support) of the source waveform and the
  source-to-receiver distance are assumed known. A least squares
  formulation of this problem exhibits the ``cycle-skipping''
  behaviour observed in field scale problems of this type, with many
  local minima differing greatly from the global minimizer. An
  extended formulation, dropping the support constraint on the source
  waveform in favor of a weighted quadratic penalty, eliminates this
  misbehaviour. With proper choice of the weight operator, the
  velocity component at {\em any} local minimizer of this extended
  objective function differs from the global
  minimizer of the least-squares formulation by less than a linear
  combination of the source waveform support radius and data
  noise-to-signal ratio.

@@

\section{Introduction}
Inverse problems based on wave equations (acoustic, elastic,
Maxwell's...) are commonplace in geophysics, nondestructive materials
testing, medical imaging, and other areas of science and engineering
in which wave motion plays an important role. These problems may be
formulated as nonlinear least squares problems, requiring optimization
of mean square misfit between observed and predicted data. Iterative
local numerical optimization methods (relatives of Newton's method)
are feasible solution approaches for the computationally large
variants of these nonlinear least squares problems that occur in
seismology and medical imaging. However, local optimization produces
only approximate stationary points, and these exhibit a tendency to
stagnate at physically irrelevant solutions, under typical conditions
of data acquisition. This ``cycle-skipping''
pathology can sometimes be overcome through use of information other than
measured wave data, however it remains a serious obstacle to effective
solution of field-scale inverse problems in wave propagation
\cite[]{GauTarVir:86,VirieuxOperto:09,Fichtner:10,Plessix:10,Schuster:17}.

Since numerical feasibility limits practical optimization approaches
to search for stationary points, it is natural to explore alternatives
to nonlinear least squares formulations for which approximate
stationary points might perforce be acceptable solutions. Many such
alternatives have been suggested and applied in both synthetic and
field data tests, in some cases with excellent results (see
\cite{HuangNammourSymesDollizal:SEG19},
\cite{PladysBrossierLiMetivier:GEO21} for recent reviews). However
numerical tests are inevitably ``a look in the rear view mirror'',
with no guarantee that the next example will not behave quite
differently. See \cite{Symes:2020} for an example of such an unpleasant
surprise. Theoretical guarantees that stationary points are
satisfactory solutions of specific inverse problems, in some specific
sense and under specified conditions, are almost entirely absent from
the literature.

This paper provides such a theoretical guarantee for a modification of
least-squares data fitting via {\em modeling operator extension},
applied to a very simple example: recovery of the (spatially
homogeneous) wave speed of an acoustic material model, together with
the waveform of an isotropic point radiator (``wavelet'') with
specified support, from the time history (``trace'') of the resulting
pressure field at a single remote location. This is perhaps the
simplest inverse wave propagation problem that exhibits
cycle-skipping. While far too simple to have immediate
practical application, it is a special case, or subproblem, of many
field-scale inverse problems in industrial and academic seismology. It
is also routinely used to illustrate cycle-skipping (see for example
\cite{VirieuxOperto:09}, Figure 7). Modeling operator extension ideas underlie
widely used techniques in seismic data processing, and have been
applied to inversion for several decades
\cite[]{geoprosp:2008}. The particular approach used in this work
is an example of {\em source extension}.
\cite{HuangNammourSymesDollizal:SEG19} review recent source extension
contributions, some of which are closely related to the approach to
single-trace transmission inversion studied here.

Model extension is only one of several ideas proposed to remedy
cycle-skipping. For instance, measuring data misfit with versions of the Wasserstein
metric from optimal transport theory, rather than the $L^2$ norm, also
shows promise in mitigating cycle-skipping
\cite[]{Metivier:GEO18,EngquistYang:GEO18,Ramos-Martinez:SEG18,Wang:SEG19,EngquistYang:CPAM21}. \cite{MahankaliYang:21}
have established conditions under which the Wasserstein misfit is
convex in a few parameters of simple models. Aside from the present
paper, their work is one of the very few to give a complete mathematical
justification (albeit also in very
restricted circumstances) for application of local optimization to a nonlinear
least squares alternative.

The next section precisely defines the single-trace transmission
inverse problem, and the following section confirms the
cycle-skipping behaviour of the obvious nonlinear least squares
formulation, that is, that stationary points exist at arbitrarily
large distance from any global minimizer (Theorem \ref{thm:fwi}). The
modeling operator extension used here simply consists in dropping the
support constraint, thus allowing fit to any data with any wavespeed,
by suitable choice of wavelet. The link between wave speed and data is
re-established by supplementing the mean-square data misfit function
with a suitable quadratic penalty on nonzero values of the wavelet far
from the specified support. I show that any stationary point of the
resulting penalty function deviates from the global solution of the
original, constrained nonlinear least squares problem by an amount
proportional to a combination of the support diameter and the relative
data error, and this bound is sharp
(Theorems \ref{thm:rampreallygood} and \ref{thm:mnoiseres}). A
solution of the original inverse problem (with wavelet support
constraint) is recovered by truncating the wavelet obtained in the
course of minimizing the penalty function (Theorem
\ref{thm:ipnoisesuf}).

The elementary
analysis developed here leans on the special features of the problem
under study to achieve sharper results than will likely be possible in
more general contexts. The Appendix recasts some of these
constructions in a form that applies, at least formally, to more
prototypical, larger-scale problems
\cite[]{Symes:IPTA14,tenKroode:IPTA14}, and suggests a route to
achieving a similar, mathematically complete understanding of
extension-based approaches to applied wave propagation inverse
problems.

\section{An Acoustic transmission Inverse Problem}
According to linear acoustics, the causal pressure field $p(\bx,t)$ due to an
isotropic point radiator at $\bx_s \in \bR^3$ with time-varying
intensity (``wavelet'') $w(t)$, propagating in an elastic fluid with
slowness (reciprocal velocity) $m$ and density $1$ (in appropriate
units), is the solution of the the initial value problem for the wave equation \cite[]{Frie:58}:
\begin{eqnarray}
  \label{eqn:awe}
  \left(m^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
  p(\bx,t)&=&0, t\ll 0.
\end{eqnarray}
The solution is well-known, see for instance
\cite{CourHil:62}, Chapter VI, section 12, equation 47:
\begin{equation}
  \label{eqn:homsol}
  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right),
\end{equation}
This trace of $p$ at $\bx_r \in \bR^3$ can be viewed as the result of
applying an m-dependent linear operator $F[m]$ to the wavelet $w$:
\begin{equation}
\label{eqn:mod}
F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right) 
\end{equation}
The slowness $m$ must be positive, as follows from basic acoustics,
and in fact reside in a range characteristic of the
material model: for crustal rock, a reasonable choice would be
$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km. The pressure trace
$p(\bx_r,\cdot)$ is square-integrable if and only if the same is true
of the wavelet $w$. Since the square-integral of the pressure trace is
proportional to the accumulated energy transferred from the fluid to
the sensor \cite[]{SantosaSymes:00}, assume that $w \in L^2(\bR)$.

Natural choices for domain and
range of $F$ are thus
\begin{itemize}
\item $M=(m_{\rm min}, m_{\rm max}),\,0 < m_{\rm min} \le m_{\rm
    max}$;
\item $W = L^2(\bR)$;
\item $D=L^2([t_{\rm min},t_{\rm max}]),\, t_{\rm min}<t_{\rm max}$;
\item $F: M \times W \rightarrow D$ as specified in \ref{eqn:mod}.
\end{itemize}
It is immediately evident from these choices and from the definition
\ref{eqn:mod} that
\begin{equation}
  \label{eqn:mapprop}
  \mbox{for }m \in M, F[m] \mbox{ is bounded, and }\|F[m]\| =
  \frac{1}{4\pi r}.
\end{equation}

Since
all possible data lie in the range of $F[m]$ for any $m \in M$, some
restriction of the domain of $F$ is necessary in order that fitting
the data constrain $m$. The constraint employed in this work is the specification
of a maxium support radius $\lambda_{\rm max} >0$ (see the companion
paper \cite{SymesChenMinkoff:21} for a justification of this choice).

For
$\lambda \in (0,\lambda_{\rm max}]$, define
\begin{itemize}
\item $\lW = \{w \in W:
  \mbox{supp }w \subset [-\lambda,\lambda]\}$;
\item $\lF = F|_{M \times \lW}$.
\end{itemize}

@@

\noindent{\bf Remark:} The constraint $\epsilon < 1$ imposed on the
target noise level eliminates the obvious choice $(m,0)$, which
satisfies the data misfit constraint for any $m \in M$ if $\epsilon
\ge 1$.

@@

@@

\section{Full Waveform Inversion}
While $F$ is surjective, as noted above, it is
very far from injective. On the other hand, under a constraint that
will be assumed throughout, $\lF[m]$ is injective for each $m \in M$ (in fact, $4 \pi r
\lF[m]$ is an isometry):
\begin{proposition}
  \label{thm:fullrec}
  Suppose that 
  \begin{equation}
    \label{eqn:fullrec}
    [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
    \subset [t_{\rm min},t_{\rm max}].
  \end{equation}
  Then $\lF[m]$ is coercive for every $m \in M, \lambda \in
  (0,\lambda_{\rm max}]$.
\end{proposition}

@@

\begin{proof} From the definition \ref{eqn:mod},
\[
 \lerr[m,w;d] =  \frac{1}{32\pi^2
    r^2\|d\|^2}\int\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
\]
Since $w_*, w$ vanish for $|t|>\lambda$,
$\lF[m_*]w_*(t)$ vanishes if $|t-m_*r|>\lambda$ and $\lF[m]w$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
either $\lF[m]w(t)=0$ or $\lF[m_*]w_*=0$. Therefore $\lF[m]w$ and
$\lF[m_*]w_*$ are orthogonal in the sense of the $L^2$ inner product
$\langle \cdot,\cdot \rangle_D$ on $D$:
\begin{equation}
  \label{eqn:ortho}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
  F[m_*]w_*\rangle_D = 0
\end{equation}
But $d = \lF[m_*]w_*$, so this is the same as saying that $d$ is
orthogonal to $F[m]w$. So conclude after a minor manipulation that
\[
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
\]
\begin{equation}
  \label{eqn:iso}
  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
\end{equation}
That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
$m$, and its minimum over $w$ is attained for $w=0$
\end{proof}

Therefore local minimizers of $\lerr$ abound, as far as you like from the
global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
$e$ gives no useful information whatever about constructive search
directions, and descent-based optimization tends to fail if the
initial estimate $m_0$ is in error by more than $2\lambda/r$. In fact the actual behaviour of FWI itererations is worse
(failure if $m_0$ is in error by ``half a wavelength''), as follows
from a more refined analysis of the cycle-skipping local behaviour of $\lerr$ near its
global minimizer.

@@

The main theoretical device used in the proofs of our main results on 
extended inversion is Variable Projection reduction of the penalty objective $\Ja$
(equation \ref{eqn:jdef}) to a function $\tJa$ of 
$m$ alone, by minimization over $w$:
\begin{equation}
  \label{eqn:redexp}
  \tJa[m;d] = \inf_w \Ja[m,w;d]
\end{equation}
A minimizer $w$ on the right-hand side of definition
\ref{eqn:redexp} must solve the {\em normal equation}
\begin{equation}
  \label{eqn:norm}
  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
\end{equation}
(here and in the following, the superscript $T$ signifies adjoint in
the sense of domain and range Hilbert structures).

@@

With these choices, the normal equation \ref{eqn:norm1} becomes
\begin{equation}
\label{eqn:norm2}
\left(\frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
      t_{\rm max}-mr]} + \alpha^2a^2\right)w= F[m]^Td.
\end{equation}
Evidently, $F^TF$ (the first summand in equation \ref{eqn:norm2}) is
not coercive. If $A$ is to yield small output for input localized near
$t=0$, as it must to play its desired role in replacing the hard
support constraint, then $A^TA$ must not be coercive either, that is,
its spectrum must have zero as its infimum. Therefore an additional
condition on $A$ is required to enable application of the Lax-Milgram
constrruction to the normal
equation \ref{eqn:norm}. This condition is an hypothesis of

\begin{proposition}
  \label{thm:norminvexp}
  Assume the conditions \ref{eqn:mod}, \ref{eqn:fullrec},
  \ref{eqn:annmult}. Also assume that $\lambda \in (0,\lambda_{\rm
    max}], \alpha > 0,$ and that  $C>0$ exists so that $a \in L^{\infty}(\bR)$
  mentioned in equation \ref{eqn:annmult} satisfies the condition
\begin{equation}
    \label{eqn:abnd} 
    a \ge 0; \, a(t) \ge C\mbox{ for }|t| \ge \lambda_{\rm max}.
  \end{equation}
%  \begin{equation}
%   \label{eqn:abnd}
%    |t| > \lambda \Rightarrow a(t) \ge C,
%  \end{equation}
%  in which $C>0$ may depend on $\lambda$.
  Then
  \begin{itemize}
  \item[1. ]the normal operator $F[m]^TF[m] + \alpha^2A^TA$ is
    invertible for any $m \in M$, $\alpha > 0$;
  \item[2. ]the solution $\aw[m;d]\in W$ of the normal equation
    \ref{eqn:norm} is given by
    \begin{equation}
      \label{eqn:normsol}
      \aw[m;d](t) = \left\{
        \begin{array}{c}
          \left(\frac{1}{(4\pi r)^2} + \alpha^2
          a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
          min}-mr, t_{\rm max}-mr];\\
          0, \mbox{ else;}
        \end{array}
      \right.
    \end{equation}
  \item[3. ]if in addition $d=F[m_*]w_*, w_* \in \lW$ is noise-free, as in equation
    \ref{eqn:defdatanonoise},
    \begin{equation}
      \label{eqn:solnonoise}
      \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right).
    \end{equation}
  \end{itemize}
\end{proposition}

@@

If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
were differentiable, then ``local minimizer'' in the conclusion of the
preceding theorem could be replaced by ``stationary point''. However,
for the problem addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is
  not} differentiable without added smoothness constraints on $w$,
whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
penalty operator $A$.
This conclusion follows from properties of the modeling operator $F$
shared with many other inverse problems in wave propagation, as
explained in the Appendix. Here, I derive differentiability from explicit
expressions for $\tJa$ and its components.
